A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-16 analysis/paper/paper_blind.Rmd
submitting to AAP...

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author:
  - "Ben Marwick"
  - "Suzanne E. Pilaar Birch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Data are the building blocks of research, and access to data is necessary to fully understand and extend the work of others (Fienberg et al 1985). Yet archaeological data are often not available for a variety of reasons, including a perceived lack of credit for making data available. The goal of this paper is to present a standard for the scholarly citation of archaeological data, akin to the way we cite publications. The need for this has been motivated by Huggert’s (2017) findings of low levels of data citation and reuse in archaeology, and Clarke’s (2015) analysis that digital data preservation and management systems are under-developed in archaeology, despite our obligations as stewards of cultural heritage. A standard for data citation is important because it gives credit to the authors or compilers of the data, and so accords them visibility and recognition for their scholarly contribution in a way that is valued broadly in the discipline. A citation standard is also important for tracking the use of the data, and generating metrics about the impact of the dataset on the discipline. Understanding the use of resources such as datasets is important for allocating resources within the discipline and identifying novel research directions.

@@

"The term “data” as used in this document is meant to be broadly inclusive. In addition to digital manifestations of literature (including text, sound, still images, moving images, models, games, or simulations), it refers as well to forms of data and databases that generally require the assistance of computational machinery and software in order to be useful, such as various types of laboratory data including spectrographic, genomic sequencing, and electron microscopy data; observational data, such as remote sensing, geospatial, and socioeconomic data; and other forms of data either generated or compiled, by humans or machines." (Uhlir & Cohen, 2011, cited in Borgman 2015)

```{r summary-of-defs-of-data}
tbl <- readr::read_csv("../figures/tbl_definitions_of_data.csv")
knitr::kable(tbl, caption = "A selection of defintions of research data")
```

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different requirements for access, archiving, and so on. For many archaeologists, data are the contents of the files we use to generate visualisations, summary tables, and qualitative statements that appear in, and support the claims we make in, our scholarly publications and reports. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." This is a highly pragmatic approach to defining data, but is also consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependant on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). It implies that the answer to the question of what counts as data is mostly emergent from our professional interactions, such as the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email.  Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how to these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). Archaeologists can be found all over this spectrum because of the diversity of archaeological research, however, our view is that archaeology is, for the most part, currently a restricted data-sharing field. There are few standardised and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

Even though there is a wide spectrum of data sharing behaviours, from intensive to restrictive, overall rates of data sharing are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA's Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a 'poor availability of data'.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2017), we include the entire R code used for all the analysis and visualizations contained in this paper in our compendium at http://dx.doi.org/10.17605/OSF.IO/KSRUZ. Also in this version-controlled compendium are the raw data for all the tests reported here. All of the figures and quantitative results presented here can be independently reproduced with the code and data in this repository. In our compendium our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2017).

@@

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and recieved `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We recieved `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research." Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promises to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has been made their data openly available.

We see similar rates of sharing in response to email requests among archaeologist to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses suggests two key reasons why author are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data, and that their data is not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas. The absence of patterns in the date of the paper, student status of the author and other variables may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

comp_perc <- data_type %>% 
  filter(`data type?`  == "compositional") %>% 
  pull(perc)

@@

dna_perc <- data_type %>% 
  filter(`data type?`  == "DNA sequence") %>% 
  pull(perc)

@@

dates_perc <- data_type %>% 
  filter(grepl("dates", `data type?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(perc)

@@

xlsx_n <- file_type %>% 
  filter(grepl("XLSX", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(n)

xlsx_perc <- file_type %>% 
  filter(grepl("XLSX", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(perc)

@@

csv_perc <- file_type %>% 
  filter(grepl("CSV", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(perc)

@@

doc_perc <- file_type %>% 
  filter(grepl("doc|DOC", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(perc)

@@

script_perc <- software %>% 
  filter(grepl("^R$|.*R$|Python|MATLAB|OxCal", `Analysis software?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  pull(perc)

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` articles published during 2016-2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviours. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, a open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that “All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article”. The data availability rate of 53% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instrumentes rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organise (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognise code for data analysis as a primary research object to share with the research community.

@@

```{r datacite-temporal-summary, fig.cap = "Change over time in archaeology datasets known to DataCite. A. Counts of datasets per year by the major data centres, red lines indicate major increases in the number of datasets in repositories, Note that the date on the horizontal axis indicates when the data was collected, not when it was deposited in the data centre, B. Temporal trends in the number of DOIs at individual data centres."}
knitr::include_graphics("../figures/datacite_temporal_plots.png")
```

@@

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, file deposition started and steadily increased during 1990-2008, as personal computers become more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This relates to the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS within two years of completion of their projects (Keers et al. 2011). Anothe spike occurs in 2012 when tDAR had a large increate in deposits, likley due to them waiving their deposit fees for that year (Ellison 2012). After 2009 the volume of deposits declined in many repositories, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

Of the `r prettyNum(sum(rights_summary$nn), big.mark = ",")` records with copyright information, the majority are released under the ADS conditions of use (n = `r prettyNum(rights_ads_n, big.mark = ",")`, `r rights_ads_perc`%), which is broadly similar to the CC-BY-NC (Moore and Richards 2015). The next most frequently used condition is the EU Open Access licence (n = `r prettyNum(rights_eu_n, big.mark = ",")`, `r rights_eu_perc`%), similar to the CC-BY licence. Creative Commons licenses such as CC-BY, CC-0 and others are used by less than 3% of records.

Many of the records are grey literature such as unpublished reports, produced by commercial consultants or research projects. Of the `r prettyNum(number_of_records_with_file_type, big.mark = ",")` (`r perc_of_records_with_file_type`%) records that include information about the file types in the deposit, `r perc_only_pdf`% (n = `r prettyNum(n_only_pdf, big.mark = ",")`) contain only PDFs. Structured data files such as spreadsheets, databases or shapefiles are found either alone or together with other file types (typically PDFs) in `r perc_structured_data_file`% (n = `r prettyNum(n_structured_data_file, big.mark = ",")`) records. The highest proportion of records with structured data files are found in the DANS-EASY repository (n = `r prettyNum(n_struc_in_dans, big.mark = ",")`, `r perc_struc_in_dans`%), followed by the ADS with `r perc_struc_in_ads`% (n = `r n_struc_in_ads`). Note that file format information is not present for all records in this sample, and some of the smaller repositories do not supply this information to DataCite at all, so it is possible that the true proportion of structured files in these records is higher than obsevered here.

From this pilot study of repository data we see that policy mandates in the UK and the Netherlands have been effective in populating repositories with archaeological data. With the 2012 spike in records deposited in tDAR we also see that archaeologists are sensitive to the price of sharing data. Nearly all records are made available under conditions that permit others to use the data while requiring that the original author of the data receive credit in reuse. In most cases this is because the repository hosting the data requires the author to choose one of these permissive licenses as a condition of making the deposit. However, although many items are _able_ to be reused because of the licenses, then ease of reuse of these items is low. Only a small fraction of the items deposited are machine-readable structured data files. The majority of items are PDFs, which require extensive manual handling to extract tablular data for convenient reuse and combination with other datasets.

@@

```{r datacite-summary, fig.cap = "Distributions of data centers, languages (en = English, nl = Dutch, swe = Swedish, de = German, it = Italian) and licenses (see text for details) among archaeology datasets known to DataCite. Horizontal axis is the number of DOIs."}
knitr::include_graphics("../figures/datacite_plots.png")
```

```{r dates, eval = TRUE}
# why so many in 2009? http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0078080 it's the Dutch

@@

The Joint Declaration of Data Citation Principles is the current standard in may disciplines for data citation within any research object (Data Citation Synthesis Group 2014). The result of these principles is a citation formation as follows (the exact order of the items may vary slightly across different referencing styles):

@@

"This paper uses data from the Open Science Framework data repository at http://doi.org/10.17605/OSF.IO/RTZTH (Marwick 2017), first described in Marwick et al. 2017."

>>>>

---
title: "A Standard for the Scholarly Citation of Archaeological Data"
author:
  - ""
  - ""
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
    reference_docx: ../templates/template.docx
    fig_caption: yes

abstract: |
  "How do archaeologists share their research data, if at all? We review what data are, according to current influential definitions, and previous work on the benefits, costs and norms of data sharing in the sciences broadly. To understand data sharing in archaeology, we present the results of three pilot studies: requests for data by email; review of data availability in published articles, and analysis of archaeological datasets deposited in repositories. We find that archaeologists are often willing to share, but discipline-wide sharing is patchy and ad hoc. Legislation and mandates are effective at increasing data-sharing, but editorial policies at journals lack adequate enforcement. Although most of data available at repositories are licensed to enable flexible reuse, only a small proportion of the data are stored in structured formats for easy reuse. We present some suggestions for improving the state of date sharing in archaeology, among these is a standard for citing data sets to ensure that researchers making their data publicly available receive appropriate credit."
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Data are the building blocks of research, and access to data is necessary to fully understand and extend the work of others (Fienberg et al 1985). Yet archaeological data are often not available for a variety of reasons, including a perceived lack of credit for making data available. The goal of this paper is to present a standard for the scholarly citation of archaeological data, akin to the way we cite publications. The need for this has been motivated by Huggert’s (2017) findings of low levels of data citation and reuse in archaeology, and Clarke’s (2015) analysis that digital data preservation and management systems are under-developed in archaeology, despite our obligations as stewards of cultural heritage. A standard for data citation is important because it gives credit to the authors or compilers of the data, and so accords them visibility and recognition for their scholarly contribution in a way that is valued broadly in the discipline. A citation standard is also important for tracking the use of the data, and generating metrics about the impact of the dataset on the discipline. Understanding the use of resources such as datasets is important for allocating resources within archaeology and identifying novel research directions.

@@

"The term "data" as used in this document is meant to be broadly inclusive. In addition to digital manifestations of literature (including text, sound, still images, moving images, models, games, or simulations), it refers as well to forms of data and databases that generally require the assistance of computational machinery and software in order to be useful, such as various types of laboratory data including spectrographic, genomic sequencing, and electron microscopy data; observational data, such as remote sensing, geospatial, and socioeconomic data; and other forms of data either generated or compiled, by humans or machines." (Uhlir & Cohen, 2011, cited in Borgman 2015)

```{r summary-of-defs-of-data}
tbl <- readr::read_csv("../figures/tbl_definitions_of_data.csv")
knitr::kable(tbl, caption = "A selection of definitions of research data")
```

This is a useful definition because it emphasizes data as digital manifestations, and so suitable for transmission, storage and retrieval using the internet. Physical objects (such as samples, collections, hard copies of notes, etc.) are data also, but they are out of scope of this discussion because of the different requirements for access, archiving, and so on. For many archaeologists, data are the contents of the files we use to generate visualisations, summary tables, and qualitative statements that appear in, and support the claims we make in, our scholarly publications and reports. The categories shown in Table \@ref(tab:types-of-data) are useful as a rough guide to what data can be, although this table is not comprehensive, and does not represent the full diversity of data used by archaeologists. Ultimately, data are whatever the community decides, as the The National Science Foundation notes in their commentary on Data Management Plans: "What constitutes such data will be determined by the community of interest through the process of peer review and program management. This may include, but is not limited to: data, publications, samples, physical collections, software and models." This is a highly pragmatic approach to defining data, but is also consistent with the relational account of data in philosophy of science scholarship, which holds that what data are is dependent on who uses it, how and for what purpose (Leonelli 2015; Daston and Galison 1992). It implies that the answer to the question of what counts as data is mostly emergent from our professional interactions, such as the materials we request when we do peer review of manuscripts and grant proposals, or when we make an informal request for more information by email.

Pragmatically in our own work, we consider that the main data suitable for sharing as the contents of the computer files (e.g. text, raw structured data, and images) that we analyse or otherwise use to create the figures, tables and quantitative results that we include in our reports and publications. The selection omits materials that could also be considered data accoring to the defintions noted above (e.g. field notes), but we feel our selection represents the most information-dense and meaningufl data records that are useful for others wanting to assess the reliability of our work, or resuse our findings in their own work. Our citation standard imposes no constraints on what data can be, but we propose a few minimal qualities for how data should be organised that will help ground our discussion. Table \@ref(tab:desirable) summarises some desirable minimum qualities for archaeological data that optimise its discovery and reuse.

@@

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how do these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). This intensive-restricted spectrum also maps on to Sawyer's (2008) analysis of data sharing in data-rich fields compared to data-poor fields. Data rich fields, such as astronomy, oceanography, and ecology, are have three common characterics: pooling and sharing data resources is expected; the cumulative nature of work that builds on common data and common methods reduces the likelihood of alternative data sources and other methods being common; and the number of theoretical choices is limited because there is enough data to adequately test competing theories. Data-poor fields have three common characteristics: data are rare, or hard to get, or both; the types of data available often dictate the methods taken; and theories are relatively easy to generate and more difficult to validate. In data-poor fields, researchers often invest substantial time and resources on gathering their own data. The getting of data, and the quality of the data collected, is often a key determinant of the status of a researcher, leading to hoarding of data as an asset to be traded for professional benefit.

Archaeologists can be found all over this spectrum because of the diversity of archaeological research. However, our observations of data sharing in archaeology, and the nature of archaeological research generally, is that it is, for the most part, currently a restricted data-sharing and a data-poor field. There are few standardised and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

Even though there is a wide spectrum of data sharing behaviours, overall rates of data sharing by individual researchers are very low in many fields. Wicherts et al. (2006) emailed authors of 141 empirical articles recently published by the American Psychological Association (APA) journals to request a copy of the original data reported in the articles. They received 64 of the 249 datasets they requested, resulting in a data sharing rate of about 26%, despite all the authors of the articles in their study being bound to the APA's Ethical Principles, which state authors must ‘not withhold the data on which their conclusions are based from other competent professionals who seek to verify the substantive claims’ (American Psychological Association, 2001, p. 396). Valpaemel et al. (2015) similarly requested data from 394 papers, published in all issues of four APA journals in 2012 and found that 38% of authors responded with their data. They concluded that although the data sharing rate increased as compared to the study by Wicherts et al., their findings are ‘worrisome’ and indicate a 'poor availability of data'.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2017), we include the entire R code used for all the analysis and visualizations contained in this paper in our compendium at http://dx.doi.org/10.17605/OSF.IO/KSRUZ. Also in this version-controlled compendium are the raw data for all the tests reported here. All of the figures and quantitative results presented here can be independently reproduced with the code and data in this repository. In our compendium our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2017).

@@

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and received `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We received `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing.

Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014). Our finding is not unique - Chambers' (2017) examined 50 PLOS One papers on brain images published in November 2014 - May 2015 and found that only 38% of these papers archived their data in an open repository. Similarly, Nuijten et al. (2017) found that of 462 PLOS One articles that promised data availability, 29% of these did not have any available data. Unfulfilled promises to share data are not unique to PLOS One (Kidwell et al. 2016), but these studies highlight the need to resolve ambiguity among authors about what counts as data, and for editors and peer reviwers to more critically assess whether an author has made their data openly available.

We see similar rates of sharing in response to email requests among archaeologists to what have been described in other disciplines (Table \@ref(tab:sharing-rates), Tenopir et al. 2015; Vanpaemel et al. 2015). However, our result may overestimate the true rate because our sample did not include datasets that authors might not be able or willing to share because they contain sensitive location or cultural information. Analysis of the qualitative responses suggests two key reasons why authors are reluctant to share: they are fearful that they will lose the opportunity to produce further publications from those data (a typical concern of researchers in data-poor fields), and that their data are not in a form suitable for sharing and the task of organising those data is not a high priority. These reasons not to share have also been documented in other research areas, noted above. The absence of patterns in the date of the paper, student status of the author and other variables may relate to more to our small sample size than a real absence of relationships in data sharing behaviour.

@@

comp_perc <- data_type %>% 
  filter(`data type?`  == "compositional") %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

dna_perc <- data_type %>% 
  filter(`data type?`  == "DNA sequence") %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

dates_perc <- data_type %>% 
  filter(grepl("dates", `data type?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

xlsx_n <- file_type %>% 
  filter(grepl("XLSX", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

xlsx_perc <- file_type %>% 
  filter(grepl("XLSX", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

csv_perc <- file_type %>% 
  filter(grepl("CSV", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

doc_perc <- file_type %>% 
  filter(grepl("doc|DOC", `What file format?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

script_perc <- software %>% 
  filter(grepl("^R$|.*R$|Python|MATLAB|OxCal", `Analysis software?`)) %>% 
  summarise(n = sum(n),
            perc = sum(perc)) %>% 
  mutate(perc = round(perc, 0)) %>% 
  pull(perc)

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` articles published during 2016-2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviours. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

The _Journal of Archaeological Science_ has had a 'data disclosure' policy since at least 2013 (Editors, 2013) which states that “All data relating to the article must be made available in Supplementary files or deposited in external repositories and linked to within the article”. The data availability rate of 53% reported in our pilot study reflects either weak enforcement of this policy or an incomplete understanding among editors and authors of how to interpret and implement this policy. The prominence of compositional and dating data among the shared data types suggests a least effort strategy by authors sharing data that do not require extensive cleaning and tidying (because the data are generated by instruments rather than entered by hand, such as artefact attribute data). The size of compositional and dating datasets means that they mostly fit easily into tables in the text or supplementary files, and do not require much effort to organise (e.g. very little metadata is required for others to use these data). The small proportion of papers in our sample using or sharing a scripting language suggests that unlike other fields (Eglen et al. 2017), archaeologists do not appear to recognise code for data analysis as a primary research object to share with the research community.

@@

```{r datacite-temporal-summary, fig.cap = "Change over time in archaeology datasets known to DataCite. A. Counts of datasets per year by the major data centres, red lines indicate major increases in the number of datasets in repositories. Note that the date on the horizontal axis indicates when the data was collected, not when it was deposited in the data centre, B. Temporal trends in the number of DOIs at individual data centres."}
knitr::include_graphics("../figures/datacite_temporal_plots.png")
```

@@

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, file deposition started and steadily increased during 1990-2008, as personal computers became more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This relates to the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS within two years of completion of their projects (Keers et al. 2011). Another spike occurs in 2012 when tDAR had a large increase in deposits, likely due to them waiving their deposit fees for that year (Ellison 2012). After 2009 the volume of deposits declined in many repositories, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

Of the `r prettyNum(sum(rights_summary$nn), big.mark = ",")` records with copyright information, the majority are released under the ADS conditions of use (n = `r prettyNum(rights_ads_n, big.mark = ",")`, `r rights_ads_perc`%), which is broadly similar to the CC-BY-NC (Moore and Richards 2015). The next most frequently used condition is the EU Open Access license (n = `r prettyNum(rights_eu_n, big.mark = ",")`, `r rights_eu_perc`%), similar to the CC-BY license. Creative Commons licenses such as CC-BY, CC-0 and others are used by less than 3% of records.

Many of the records are grey literature such as unpublished reports, produced by commercial consultants or research projects. Of the `r prettyNum(number_of_records_with_file_type, big.mark = ",")` (`r perc_of_records_with_file_type`%) records that include information about the file types in the deposit, `r perc_only_pdf`% (n = `r prettyNum(n_only_pdf, big.mark = ",")`) contain only PDFs. Structured data files such as spreadsheets, databases or shapefiles are found either alone or together with other file types (typically PDFs) in `r perc_structured_data_file`% (n = `r prettyNum(n_structured_data_file, big.mark = ",")`) records. The highest proportion of records with structured data files are found in the DANS-EASY repository (n = `r prettyNum(n_struc_in_dans, big.mark = ",")`, `r perc_struc_in_dans`%), followed by the ADS with `r perc_struc_in_ads`% (n = `r n_struc_in_ads`). Note that file format information is not present for all records in this sample, and some of the smaller repositories do not supply this information to DataCite at all, so it is possible that the true proportion of structured files in these records is higher than observed here.

From this pilot study of repository data we see that policy mandates in the UK and the Netherlands have been effective in populating repositories with archaeological data. With the 2012 spike in records deposited in tDAR we also see that archaeologists are sensitive to the price of sharing data. Nearly all records are made available under conditions that permit others to use the data while requiring that the original author of the data receive credit in reuse. In most cases this is because the repository hosting the data requires the author to choose one of these permissive licenses as a condition of making the deposit. However, although many items are _able_ to be reused because of the licenses, the ease of reuse of these items is low. Only a small fraction of the items deposited are machine-readable structured data files. The majority of items are PDFs, which require extensive manual handling to extract tabular data for convenient reuse and combination with other datasets.

@@

```{r datacite-summary, fig.cap = "Distributions of data centers, languages (en = English, nl = Dutch, swe = Swedish, de = German, it = Italian) and licenses (see text for details) among archaeology datasets known to DataCite. Horizontal axis is the number of DOIs."}
knitr::include_graphics("../figures/datacite_plots.png")
```
```{r dates, eval = TRUE}
# why so many in 2009? http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0078080 it's the Dutch

@@

The Joint Declaration of Data Citation Principles is the current standard in many disciplines for data citation within any research object (Data Citation Synthesis Group 2014). The result of these principles is a citation formation as follows (the exact order of the items may vary slightly across different referencing styles):

@@

"This paper uses data from the Open Science Framework data repository at http://doi.org/10.17605/OSF.IO/32A87 (Marwick 2017), first described in Marwick et al. 2017."

Citing both the dataset and the publication helps to resolve ambiguity about the source of the data, and gives context to the data. Furthermore, the authorship of the data files may differ from those of the author's paper describing the data, and it is important that both authors (or sets of authors) receive credit for their work through the normal mechanisms of citation recognition.
